The Chemistry Development Kit (CDK): atom typing, rendering, molecular formulas, and substructure searching

### 2016-10-05 article.tex
The term 'open source' is so common now, that it no longer needs to be capitalized

\begin{abstract}
% Do not use inserted blank lines (ie \\) until main body of text.
\parttitle{Background}
The Chemistry Development Kit (CDK) is a 
widely used Open Source cheminformatics toolkit, providing data
structures to represent chemical concepts along with methods to
manipulate such structures and perform computations on them. The
library implements a wide variety of cheminformatics algorithms
ranging from chemical structure canonicalization to molecular
descriptor calculations and pharmacophore perception. Over the
last five years, the code based has grown significantly resulting in
many, complex interdependencies between components and poor
performance of many algorithms. 
\parttitle{Results}
We report improvements to the CDK since the 1.2 release series,
specifically addressing the increase functional complexity and poor
performance. We first summarize the addition of new, and
improvements to existing functionality that has led to significant
improvement in performance and support for important chemistry
features including atom type perception, substructure searching, molecular
fingerprints, rendering of molecules, and handling of molecular formulas.
Second, we outline how the CDK has evolved with respect to
quality control and the approaches we have adopted to ensure stability, including
a code review mechanism.
\parttitle{Conclusions}
This paper highlights the our continued efforts to provide a community
driven, Open Source
cheminformatics library, and show that such collaborative projects can exist over
extended periods of time, resulting in a high-quality and performant
library. By taking 
advantage of community support and contributions, we show 
that an Open Source cheminformatics project can act as a peer reviewed
publishing platform for scientific computing software.
\end{abstract}

@@

The Open Source cheminformatics community has made significant steps
forward recently~\cite{OBoyle2011b} as evidenced by the growing number
of tools and underlying toolkits, along with the usage of these
software components in a variety of applications.
The Chemistry Development Kit (CDK) is one of the tools developed
under the aegis of the Blue Obelisk 
and previously documented versions have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
Adoption of the CDK ranges from inclusion of CDK functionality in
wrapper platforms such as Cinfony~\cite{OBoyle2008}, incorporation
within the R environment (rcdk~\cite{Guha2007}), and as plugins for
Taverna~\cite{Truszkowski2011}, 
KNIME~\cite{Beisken2013}, Cytoscape (ChemViz2~\cite{ChemViz2}), and for
Microsoft Excel (LICSS~\cite{Lawson2012}).
In contrast to scenarios that have made CDK functionality available in
larger systems, a number of projects have employed the CDK as a
general cheminformatics toolkit. Examples include 
jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
PaDEL~\cite{yap2011padel}, ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016},
WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, and the IUPHAR/BPS Guide to Pharmacology database~\cite{Southan2016}.
A number of such tools were initially developed using older versions
of the CDK and are updated to new releases as they are made
available. Examples include Bioclipse~\cite{spjuth2007bioclipse,
spjuth2009bioclipse} and
AMBIT~\cite{jeliazkova2011ambit,jeliazkova2011ambitsmarts,kochev2013ambit}. The
CDK has also played a role in a number of chemical studies, such as finding
the maximally bridging rings in chemical structures~\cite{Marth2015}, and
defining fingerprints for organic reactions~\cite{Segler2016}.

@@

The growth of Open Source software over the last 10 years [REF] is evidence
of the ability of communities of developers to develop systems and
processes that lead to high quality software systems for long term
use. The CDK is no different. The adoption of automatic build systems
and quality control methodologies such as unit testing, automated
source code validation, and peer review by fellow developers have
greatly improved the stability of the library, While it has slowed
development somewhat, it has allowed for cleaning up interdependencies
between modules of code, and importantly, has improved the scalability
of the development model.  This has resulted in significant new
functionality in core APIs while maintaining the quality of code
depending on those core APIs.

>>>>

\begin{abstract}
% Do not use inserted blank lines (ie \\) until main body of text.
\parttitle{Background}
The Chemistry Development Kit (CDK) is a 
widely used open source cheminformatics toolkit, providing data
structures to represent chemical concepts along with methods to
manipulate such structures and perform computations on them. The
library implements a wide variety of cheminformatics algorithms
ranging from chemical structure canonicalization to molecular
descriptor calculations and pharmacophore perception. Over the
last five years, the code based has grown significantly resulting in
many, complex interdependencies between components and poor
performance of many algorithms. 
\parttitle{Results}
We report improvements to the CDK since the 1.2 release series,
specifically addressing the increase functional complexity and poor
performance. We first summarize the addition of new, and
improvements to existing functionality that has led to significant
improvement in performance and support for important chemistry
features including atom type perception, substructure searching, molecular
fingerprints, rendering of molecules, and handling of molecular formulas.
Second, we outline how the CDK has evolved with respect to
quality control and the approaches we have adopted to ensure stability, including
a code review mechanism.
\parttitle{Conclusions}
This paper highlights the our continued efforts to provide a community
driven, open source
cheminformatics library, and show that such collaborative projects can exist over
extended periods of time, resulting in a high-quality and performant
library. By taking 
advantage of community support and contributions, we show 
that an open source cheminformatics project can act as a peer reviewed
publishing platform for scientific computing software.
\end{abstract}

@@

The open source cheminformatics community has made significant steps
forward recently~\cite{OBoyle2011b} as evidenced by the growing number
of tools and underlying toolkits, along with the usage of these
software components in a variety of applications.
The Chemistry Development Kit (CDK) is one of the tools developed
under the aegis of the Blue Obelisk 
and previously documented versions have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
Adoption of the CDK ranges from inclusion of CDK functionality in
wrapper platforms such as Cinfony~\cite{OBoyle2008}, incorporation
within the R environment (rcdk~\cite{Guha2007}), and as plugins for
Taverna~\cite{Truszkowski2011}, 
KNIME~\cite{Beisken2013}, Cytoscape (ChemViz2~\cite{ChemViz2}), and for
Microsoft Excel (LICSS~\cite{Lawson2012}).
In contrast to scenarios that have made CDK functionality available in
larger systems, a number of projects have employed the CDK as a
general cheminformatics toolkit. Examples include 
jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
PaDEL~\cite{yap2011padel}, ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016},
WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, and the IUPHAR/BPS Guide to Pharmacology database~\cite{Southan2016}.
A number of such tools were initially developed using older versions
of the CDK and are updated to new releases as they are made
available. Examples include Bioclipse~\cite{spjuth2007bioclipse,
spjuth2009bioclipse} and
AMBIT~\cite{jeliazkova2011ambit,jeliazkova2011ambitsmarts,kochev2013ambit}. The
CDK has also played a role in a number of chemical studies, such as finding
the maximally bridging rings in chemical structures~\cite{Marth2015}, and
defining fingerprints for organic reactions~\cite{Segler2016}.

@@

The growth of open source software over the last 10 years [REF] is evidence
of the ability of communities of developers to develop systems and
processes that lead to high quality software systems for long term
use. The CDK is no different. The adoption of automatic build systems
and quality control methodologies such as unit testing, automated
source code validation, and peer review by fellow developers have
greatly improved the stability of the library, While it has slowed
development somewhat, it has allowed for cleaning up interdependencies
between modules of code, and importantly, has improved the scalability
of the development model.  This has resulted in significant new
functionality in core APIs while maintaining the quality of code
depending on those core APIs.
